\Large{Learning to Pivot with Adversarial Networks}

### 2016-10-31 tex/paper.tex
move fig1
abstract + intro

Many inference problems involve data generation processes that are not uniquely specified or are uncertain in some way.
In a scientific context, the presence of several plausible data generation processes is often
associated to the presence of systematic uncertainties.
Robust inference is possible if it is based on a pivot -- a quantity whose
distribution is invariant to the unknown value of the nuisance parameters that parametrize a family of data generation processes.
We introduce an adversarial training procedure to enforce the pivotal property on a predictive model, which
allows one to tune the tradeoff between power and robustness.
We demonstrate the effectiveness of this approach with a toy and real world example.

@@

We will focus on supervised learning techniques where the presence of systematic
uncertainty can be associated to the fact that the data generation process is not uniquely
specified. In other words, the lack of systematic uncertainties corresponds to the case
that the process that generates training data is unique, fully specified,
and an accurate representative of the real world data.

A common situation when systematic uncertainty is present is when the training data are not representative of
the real data that a predictive model will be applied to in practice. Several techniques for domain adaptation have
been developed to create predictive models that are more robust to this type of uncertainty.

Another common situation is that there are several plausible data generation processes. We consider the situation of a
family of data generation processes parametrized by nuisance parameters, which can be categorical or continuous. In particular, we consider the probability model $p(X,Y,Z)$, where $X$ are the data, $Y$ are the target labels, and $Z$ are the nuisance parameters. We will consider the problem of learning a predictive model for $Y$ conditional on the observed values of $X$ that is robust to uncertainty in the unknown value of $Z$.

\kcnote{Fill in outline}

In Sec.~\ref{sec:problem}  Sec.~\ref{sec:method}  Sec.~\ref{sec:theory}  Sec.~\ref{sec:toy}  Sec.~\ref{sec:hep}
 Sec.~\ref{sec:related}  Sec.~\ref{sec:conclusions} .

@@

To minimize the effects of $Z$ in the background events, we train a classifier
using  Algorithm~\ref{alg:adversarial-training} but consider the adversarial
term ${\cal L}_r$ conditioned on $Y=0$ only. Both $f$ and $r$ are neural
networks with 3 hidden layers of 64 nodes and ReLU activations, each terminated
by a single final output node with a sigmoid activation. Experiments are performed
on a subset of 150000 samples for training while AMS is evaluated on an independent test set of 5000000
samples. Both training and testing samples are weighted such
that the signal-to-background ratio is 1/10, allowing us to probe the
efficacy of the method proposed here in a background dominated
environment. Results reported below are averages over 5 runs.

@@

As  Fig.~\ref{fig:physics-ams} illustrates, without adversarial training (at
$\lambda=0$), the maximum significance peaks at $7$. By contrast, as the
independence constraint is made stronger (for $\lambda > 0$) the AMS peak moves
higher, with a maximum value around  $7.8$ for $\lambda=10$. In other words,
trading classification accuracy for independence with respect to pileup results
in a positive total net benefit in terms of statistical significance. Setting
$\lambda$ too high however (e.g. $\lambda=500$) results in a decrease of the
maximum significance, by focusing the capacity of $f$ too strongly on
independence, at the expense of classification accuracy. As demonstrated in this
example, controlling the classification versus pivot trade-off through $\lambda$
therefore gives us a principled and effective approach for maximizing
significance by desensitizing the classifier output $f(X;\theta_f)$ in the most
beneficial way.

>>>>

Many inference problems involve data generation processes that are not uniquely
specified or are uncertain in some way. In a scientific context, the presence of
several plausible data generation processes is often associated to the presence
of systematic uncertainties. Robust inference is possible if it is based on a
pivot -- a quantity whose distribution is invariant to the unknown value of the
nuisance parameters that parametrize the family of data generation processes. In
this work, we introduce an adversarial training procedure to enforce the pivotal
property on a predictive model, which allows one to tune the tradeoff between
power and robustness. We derive theoretical results proving that the proposed
procedure converges towards a predictive model that is both optimal and independent on
the nuisance parameters, and demonstrate the effectiveness of this approach with a
toy and real world examples.

@@

In this work, we focus on supervised learning techniques where the presence of
systematic uncertainty can be associated to the fact that the data generation
process is not uniquely specified. In other words, the lack of systematic
uncertainties corresponds to the (rare) case that the process that generates
training data is unique, fully specified, and an accurate representative of the
real world data.
A common situation when systematic uncertainty is present is when the training
data are not representative of the real data that a predictive model will be
applied to in practice. Several techniques for domain adaptation have been
developed to create predictive models that are more robust to this type of
uncertainty.
Another common situation is that there are several plausible data generation
processes, specificied as a family of data generation processes
parametrized by nuisance parameters, which can be categorical or continuous.

Assuming a probability model $p(X,Y,Z)$, where $X$ are the
data, $Y$ are the target labels, and $Z$ are the nuisance parameters,
we consider the problem of learning a predictive model for $Y$ conditional on the
observed values of $X$ that is robust to uncertainty in the unknown value of
$Z$. More specifically, we introduce a learning procedure for enforcing the
pivotal property on a predictive model by jointly training adversarial networks.
We derive theoretical results proving that the proposed procedure converges
towards a predictive model that is both optimal and independent on the nuisance
parameters, and demonstrate the effectiveness of this approach with a toy and
real world examples.

The remainder of the paper is structured as follows. In Sec.~\ref{sec:problem},
we first formally define the problem of learning a pivotal predictive model. In
sections \ref{sec:method} and \ref{sec:theory}, we describe how adversarial
networks can be trained to obtain predictive models that are independent on
nuisance parameters and show that the proposed procedure indeed converges
towards a model that is both optimal and robust. The  effectiveness of the
approach is then demonstrated in sections \ref{sec:toy} and \ref{sec:hep} with a
toy and real world examples. In Sec.~\ref{sec:related}, we relate our work
to close contributions from statistical inference, domain adaptation
and fairness in classification. Finally, Sec.~\ref{sec:conclusions} gathers
our concluding remarks.

@@

To minimize the effects of $Z$ in the background events, we train a classifier
using  Algorithm~\ref{alg:adversarial-training} but consider the adversarial
term ${\cal L}_r$ conditioned on $Y=0$ only, as outlined in
Sec.~\ref{sec:problem}. Both $f$ and $r$ are neural networks with 3 hidden
layers of 64 nodes and ReLU activations, each terminated by a single final
output node with a sigmoid activation. Experiments are performed on a subset of
150000 samples for training while AMS is evaluated on an independent test set of
5000000 samples. Both training and testing samples are weighted such that the
signal-to-background ratio is 1/10, allowing us to probe the efficacy of the
method proposed here in a background dominated environment. Results reported
below are averages over 5 runs.

@@

As  Fig.~\ref{fig:physics-ams} illustrates, without adversarial training (at
$\lambda=0|Z=0$ when building a classifier at the nominal value $Z=0$ only, or
at $\lambda=0$ when building a classifier on data sampled from $p(X,Y,Z)$), the
maximum significance peaks at $7$. By contrast, as the independence constraint
is made stronger (for $\lambda > 0$) the AMS peak moves higher, with a maximum
value around  $7.8$ for $\lambda=10$. In other words, trading classification
accuracy for independence with respect to pileup results in a positive total net
benefit in terms of statistical significance. Setting $\lambda$ too high however
(e.g. $\lambda=500$) results in a decrease of the maximum significance, by
focusing the capacity of $f$ too strongly on independence, at the expense of
classification accuracy. As demonstrated in this example, controlling the
classification versus pivot trade-off through $\lambda$ therefore gives us a
principled and effective approach for maximizing significance by desensitizing
the classifier output $f(X;\theta_f)$ in the most beneficial way.
